\subsection{Size}

Althought the language is already complete, it would prove useful for further
analysis to define the notion of \emph{size}, and hence the equality and order
of relations on data values. Without further a-do, we define the size of a
value to be \emph{the number of nodes in the binary tree}.

The euqality and order relations are a bit more complicated though, as that
there is seemingly no \emph{elegant} way to decide whether one arbitrary binary
tree has the same or ..

Hence, the tree \mono{0} has the value $0$, the tree \mono{0.0} has the value
$1$, and the tree \mono{0.0.0} has the value $2$ as does it's symmetrical
equivalent, \mono{(0.0).0}.

This allows us to define the, otherwise built-in, function \mono{less} in a
primitive recursive fashion as follows:

\begin{verbatim}
less 0 0 = 0
less _._ 0 = 0
less 0 _._ = 0.0
less A.B X.Y = 

less AR.AL BR.BL = or (and (less AR BR) (less AL BL))
\end{verbatim}

This definition indicates that we choose for the empty tree to represent the
value \emph{false}, and for the tree \mono{0.0} to represent the value
\emph{true}. We'll keep the definition even more generic, and let the
\emph{nonempty} tree represent the value \emph{true}, as shall become useful
when we define the higher-order function \mono{if}
(\referToSection{language-higher-order-built-ins}).

Since the values begin at $0$ and grow at the rate of $1$ ... we can define it
as syntactic sugar and use nonnegative integers where ...

In addition to defining the actual data type we need to specify how we're going
to reason about it. Specifically, the questions of equality and order of values
constructed in this manner have to be answered.

For all intents and purposes, we can let the \emph{absolute value} of such a
tree-structured value be equal to $n-1$, where $n$ is the number of leafs in
the tree. Hence, the tree \mono{0} denotes $0$, \mono{0.0} denotes 1,
\mono{0.0.0} denotes 2 and so on.

The choice of this data representation yields the following properties for the
construction and destruction operators:

\begin{lemma} Construction of value yields a value strictly greater than either
of it's constituents. Specifically, the absolute value of the new value is the
sum of the absolute values of the constituents.\end{lemma}

\begin{lemma} Destruction of a value yields a pair of values who's absolute
values are strictly less than the absolute value of the original
value.\end{lemma}

\subsection{Programs}

Programs are defined in a conventional functional context and without mutual
recursion, namely:

\begin{align}
\nonterm{program}\ ::=&\ \nonterm{function}^*\ \nonterm{expression}
\end{align}

The order of the function definitions does matter wrt. pattern matching in so
far as those defined before are attempted first, if the match fails, the next
function with the same signature\footnote{In this case comprising of the name
of the function and it's arity.} is attempted.

Note, that we let the number of function definitions be zero as an
$\nonterm{expression}$ is a valid program as well. More generally, the program
can be thought of as a constant function, where the actual
$\nonterm{expression}$ simply has access to some predefined functions defined
by the function definitions in the program.

\subsection{Built-in high-order
functions}\label{section:language-higher-order-built-ins}

Although \mono{D} is initially a first-order language, we will ignore that
limitation for a bit and define a few higher-order functions to provide some
syntactical sugar to the language. Beyond the discussion in this section, these
higher-order functions should be regarded as \mono{D} built-ins.

\subsubsection{Branching}

In the following definition, the variable names \mono{true} and \mono{false}
refer to expressions to be executed in either case.

\begin{verbatim}
if 0 _ false := false
if _._ _ true := true
\end{verbatim}

As you can see, we employ the C convention that any value other than $0$ is a
``truthy'' value, and the expression \mono{true} is returned.

Although the call-by-value nature of the language does not allow for
short-circuiting the if-statements defined in such a way, this shouldn't be any
impediment to further analysis.

\subsection{Sample programs}

As an illustration of the language syntax, the following program reverses a tree:

\begin{verbatim}
reverse 0 := 0
reverse left.right := (reverse right).(reverse left)
\end{verbatim}

The following program computes the Fibonacci number \mono{n}:

\begin{verbatim}
fibonacci 0 x y := 0
fibonacci 0.0 x y := y
fibonacci n x y := fibonacci (minus n 0.0) y (add x y)
\end{verbatim}

\section{Semantics}

In the following section, the operational semantics of the language \mono{D}
are defined in terms of structured operaitonal semantics\cite{sos}.
\referToTable{sos-definitions} specifies most\footnote{The rest is discussed
further below and in \referToSection{language-semantics-memory}.} of the
syntactical elements used to define the semantic reduction rules below.

In addition to the notation specified in the table, we'll make use of
Haskell-like list comprehension when dealing with lists of elements. For
instance, $[e]$ refers to a list of expressions, and $[e'|e]$ refers to a list
of expressions that starts with the expression $e'$ and is followed by the list
of expressions, $e$. Also a bit alike Haskell, in both cases above, $e$ is used
as both a type and a variable.

\makeTable
{sos-definitions}
{Some of the syntactical elements used in the reduction rules for \mono{D}.}
{|lll|}
{\textbf{Notation}&\textbf{Description}}
{
$e$ & expression\\
$v$ & value\\
$n$ & variable name\\
$p$ & pattern\\
$0$ & the atom $0$\\
$\cdot$ & $\term{.}$\\
$\sigma$ & memory\\
$\sigma[n]$ & the value of variable $n$ in memory, returns some $v$
}

\subsection{Memory}\label{section:language-semantics-memory}

For the sake of an elegant notation, we'll define the notion of memory as a set
of stacks, one for each variable in the program. If a variable $n$ has an empty
stack, it is undefined, otherwise the value of the variable (in the present
scope) is the value at the top of the corresponding stack.

This model of memory allows us to deal with abitrary scope\footnote{Although
first-order \mono{D} can make little use of that.} in a rather elegant matter,
where we simply push a new value onto the corresponding stack when we enter a
nested scope and pop off the corresponding stacks when exiting a nested scope.
Hence, visiting a nested scope has the following operational semantics wrt.
$\sigma$:

$$\sigma\longrightarrow\sigma(n)\leftarrow v\longrightarrow\sigma$$

Following the conventions of structured operational semantics, the value of an
element, such as $\sigma$, does not change throughout a reduction rule. So in
the above example, the starting $\sigma$ is equivalent to the final $\sigma$.

\subsection{Evaluation}

\mono{D} has only one operator, namely $\term{.}$, which is a right-associative
binary operator, hence expressions are evaluated using the following triplet of
rules:

\begin{equation}
{
\left\langle e',\ \sigma\right\rangle
\longrightarrow
\left\langle e'',\ \sigma\right\rangle
}\over {
\left\langle e \cdot e',\ \sigma\right\rangle
\longrightarrow
\left\langle e \cdot e'',\ \sigma\right\rangle
}
\end{equation}

\begin{equation}
{
\left\langle e,\ \sigma\right\rangle
\longrightarrow
\left\langle e',\ \sigma\right\rangle
}\over {
\left\langle e\cdot v,\ \sigma\right\rangle
\longrightarrow
\left\langle e' \cdot v,\ \sigma\right\rangle
}
\end{equation}

\begin{equation}
\left\langle v \cdot v',\ \sigma\right\rangle
\longrightarrow
\left\langle v'',\ \sigma\right\rangle
\ \ \ \ \ \ (\text{where}\ v'' = v \cdot v')
\end{equation}

\subsubsection{Variables}

As expressions may contain variable names, we need a way to retrieve the values
of variables from memory:

\begin{equation}
\left\langle n,\ \sigma\right\rangle
\longrightarrow
\left\langle \sigma[n],\ \sigma\right\rangle
\end{equation}

\subsubsection{Application}

Function application is left-associative and has variable (constant at run
time) arity of at least one. The arity of the application depends on the
declaration that the function specifier, $f$, points to. Hence, we begin by
evaluating the function specifier itself to some expression $\lambda$ and a
pattern list $[p]$, that is, it's corresponding function declaration:

\begin{equation}
{
\left\langle f,\ \sigma\right\rangle
\longrightarrow
\left\langle \left\langle \lambda,\ [p]\right\rangle,\ \sigma\right\rangle
}\over{
\left\langle \left\langle f,\ [e]\right\rangle,\ \sigma\right\rangle
\longrightarrow
\left\langle
\left\langle\left\langle \lambda,\ [p]\right\rangle,\ [e]\right\rangle
,\ \sigma\right\rangle
}
\end{equation}

\emph{Note, in a first-order context $\lambda$ bares no special meaning,
however, the letter is carefully chosen to aid further extension of \mono{D} to
it's higher-order sibling.}

Followed by evaluation of the pattern and expression lists:

\begin{equation}
{
\left\langle p',\ \sigma\right\rangle
\longrightarrow
\left\langle p'',\ \sigma\right\rangle
\wedge
\left\langle e',\ \sigma\right\rangle
\longrightarrow
\left\langle e'',\ \sigma\right\rangle
}\over{
\left\langle p\cdot p',\ e\cdot e',\ \sigma\right\rangle
\longrightarrow
\left\langle p\cdot p'',\ e\cdot e'',\ \sigma\right\rangle
}
\end{equation}

\begin{equation}
{
\left\langle e',\ \sigma\right\rangle
\longrightarrow^*
\left\langle 0,\ \sigma\right\rangle
}\over{
\left\langle p\cdot 0,\ e\cdot e',\ \sigma\right\rangle
\longrightarrow
\left\langle p,\ e,\ \sigma\right\rangle
}
\end{equation}

\begin{equation}
{
\left\langle e',\ \sigma\right\rangle
\longrightarrow^*
\left\langle v',\ \sigma\right\rangle
}\over{
\left\langle p\cdot n,\ e\cdot e',\ \sigma\right\rangle
\longrightarrow
\left\langle p,\ e,\ \sigma(n)\leftarrow v' \right\rangle
}
\end{equation}

%\begin{equation}
%{
%\left\langle p',\ \sigma\right\rangle
%\longrightarrow
%\left\langle p'',\ \sigma\right\rangle
%\wedge
%\left\langle e',\ \sigma\right\rangle
%\longrightarrow
%\left\langle e'',\ \sigma\right\rangle
%}\over{
%\left\langle p'\cdot n,\ e'\cdot n,\ \sigma\right\rangle
%\longrightarrow
%\left\langle p''\cdot n,\ e''\cdot n,\ \sigma(n)\leftarrow v' \right\rangle
%}
%\end{equation}



We complete application by evaluating $\lambda$ with the new memory state:

\begin{equation}
\left\langle \lambda,\ \sigma([n])\leftarrow[v]\right\rangle
\longrightarrow
\left\langle v',\ \sigma\right\rangle
\end{equation}

\emph{Note, we return to the original $\sigma$ once $\lambda$ is evaluated.}

\newpage

This is small-step..

\begin{equation}
{
\left\langle n_f, [e], \sigma\right\rangle
\longrightarrow
\left\langle e_f, [p], [e], \sigma\right\rangle
\ \ \ \ \ \text{where}\ \left\langle e_f, [p] \right\rangle
\ \text{is the definition of the function with the name}\ n_f\text{.}
\over
\left\langle n_f, [e], \sigma\right\rangle
\longrightarrow
\left\langle v, \sigma\right\rangle
}
\end{equation}
