\begin{thebibliography}{99}

\bibitem{algol-bnf}

P. Naur (ed.), Revised Report on the Algorithmic Language ALGOL 60; CACM, Vol.
6, p. 1; The Computer Journal, Vol. 9, p. 349; Num. Math., Vol. 4, p. 420.
(1963); Section 1.1.

\bibitem{turing-machine}

A. M. Turing, On computable numbers with an application to the
Entscheidungsproblem; Proceedings of the London Mathematical Society,
42(2):230-265, (1936).

\bibitem{sos}

Gordon D. Plotkin, A Structural Approach to Operational Semantics, Journal of
Logic and Algebraic Programming (2004) Volume: 60-61, Issue: January,
Publisher: Citeseer, Pages: 17-139.

\bibitem{kildall}

Gary A. Kildall, A unified approach to global program optimization, Proceedings
of the 1st annual ACM SIGACT-SIGPLAN symposium on Principles of programming
languages, Boston, Massachusetts, 194-206, 1973.

\bibitem{shivers}

O. Shivers, Control flow analysis in scheme, Proceedings of the ACM SIGPLAN
1988 conference on Programming Language design and Implementation,
0-89791-269-1, Atlanta, Georgia, United States, 164-174, 1988.

\bibitem{jones-lambda-cfg}

Jones, Neil, Editors: Even, Shimon \& Kariv, Oded, Flow analysis of lambda
expressions, Automata, Languages and Programming, Lecture Notes in Computer
Science, Publisher: Springer Berlin / Heidelberg, ISBN: 978-3-540-10843-6,
114-128, Volume 115, 1981.

\bibitem{complexity-of-loop-programs}

Albert R. Meyer \& Dennis M. Ritchie, The complexity of loop programs,
Proceedings of the 1967 22nd national conference, Washington, D.C., United
States, 465-469, 1967.

\end{thebibliography}
