\chapter{Extended-BNF}\label{appendix:ebnf}

This report makes use of an extended version of the Backus-Naur form (BNF).
This appendix is provided to cover the extensions employed in the report. This
is done because there is seemingly no universally acknowledged extension,
unlike there is a universally acknowledged Backus-Naur form, namely the one
used in the ALGOL 60 Reference Manual\cite{algol-bnf}.

\section{What's in common with the original BNF}

The following parts are in-common with the original Backus-Naur form:

\makeTable{bnf}{Constructs in common with the original BNF.}
{|c|p{0.5\textwidth}|}
{\textbf{Construct} & \textbf{Description}}
{
$<\ldots>$ & A metalinguistic variable, aka. a nonterminal.\\
$::=$ & Definition symbol\\
$|$ & Alternation symbol
}

In the original BNF, everything else represents itself, aka. a terminal. This
is not preserved in this extension -- all terminals are encapsulated into
single quotes.

\section{Constructs borrowed from regular expressions.}

The use of single quotes around all terminals allows us to give characters such
as $($, $)$, $]$, $]$, $*$, $+$, and ${}^*$ special meaning, namely:

\makeTable{ebnf}{Constructs borrowed from regular expressions.}
{|c|l|}
{\textbf{Construct}&\textbf{Meaning}}
{
$(\ldots)$ & Entity group\\
$[\ldots]$ & Character group\\
$\text{-}$ & Character range\\
$*$ & $0\text{-}\infty$ repetition\\
$+$ & $1\text{-}\infty$ repetition\\
$?$ & $0\text{-}1$ repetition
}

An entity group is a shorthand for an auxiliary nonterminal declaration. This
means, for instance, that using the alternation symbol within it would mean an
alternation of entity sequences within the entity group rather than the entire
declaration that contains the entity group.

A character group may only contain single character terminals and an
alternation of the terminals is implied from their mere sequence. It is
identical to an auxiliary single character nonterminal declaration. A character
range binary operator can be used to shorten a given character group, e.g.
$[\term{a}\mathmono{-}\term{z}]$ implies the list of characters from $\term{a}$
to $\term{z}$ in the ASCII table.  Moreover, a character range is the only
operator allowed in a character group.

Applying the repetition operators to either the closing brace of an entity
group or the closing bracket of a character group has the same effect as
applying the repetition operator to their respective hypothetical auxiliary
declarations.

\section{Nonterminals as sets and conditional declarations}

Another extension to the original BNF is the ability to use nonterminals as
sets in declaration conditions. For example, if the two nonterminals,
$\nonterm{type-name}$ and $\nonterm{constructor-name}$, are both declared in
terms of the $\nonterm{literal}$ nonterminal, but type names and constructor
names should not intersect in a given program, then we can append the following
condition to one or both declarations:

$$\text{s.t.}\
\nonterm{type-name}\cap\nonterm{constructor-name}\equiv\emptyset$$

Where the shorthand s.t. stands for ``such that''. This implies that the
nonterminals $\nonterm{type-name}$ and $\nonterm{constructor-name}$ represent
the sets of character sequences that end up associated with the respective
nonterminals for any given program, and can be used in conjunction with regular
set notation.
